\section{Problem Statement}
The main goal of our thesis is to determine, whether the approximate histograms
produced by Kmerlight or KmerGenie can be used as inputs for CovEst that 
subsequently estimates the genome size. We mainly focus on preserving the 
accuracy of CovEst's estimates and we focus less on the processing time and memory consumption. 

In order to achieve our goal, we study the performance of CovEst on these approximate
histograms and we look for ways to increase CovEst's accuracy. There are two approaches
that come into mind: either we alter the histogram-counting-algorithm and the approximate
histogram to better suit the needs of CovEst, or we alter the probabilistic model of CovEst to
be able to cope with imprecise histograms. We explore both approaches. 

In chapter \ref{sec:chapter2} we study the qualitative and quantitative character
of inaccuracies introduced by Kmerlight\footnote{The analysis of KmerGenie's errors is trival
and it will be presented later in \ref{sec:kmergenie-errors}.} with aim to model or
predict the distribution of the estimates of $f_i$. We discover that Kmerlight can be modified
to produce more accurate histograms.

In chapter \ref{sec:chapter3} we analyse CovEst. We find that CovEst manages Kmerlight's and
KmerGenie's errors by design and that we are unable to further increase its accuracy on
approximated histograms. Nevertheless, we present a new method which helps CovEst with
a different type of histogram errors that might be used in the future research.
